---
title: "Reproducible Research: Peer Assessment 1"
author: Chris Coles
date: Nov 8th 2015
output: 
  html_document:
    keep_md: true
---
<<<<<<< HEAD

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.
=======
Testing
>>>>>>> origin/master

## Loading and preprocessing the data
The following code assumes that the data files have been downloaded and unzipped in the current working directory.

The data is loaded from activities.csv and the summary information is as follows:

```{r echo=TRUE, results="asis"}
activities <- read.csv(file= "activity.csv", header=TRUE)
sum_info <- xtable(summary(activities))


```

```{r via kable, echo=FALSE, results='asis'}
library(knitr)
library(xtable)
sum_info <- xtable(summary(activities))
kable(sum_info, format = "html", padding = 5, row.names=TRUE)

```

## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
